%% sections/05-results.tex

\section{Results}
\label{sec:results}

\subsection{Session Coverage and Instance Frequency}

Table~\ref{tab:coverage} reports session coverage and instance frequency by campaign.

\begin{table}[t]
\caption{NPC Arc-Completion Coverage by Campaign}
\label{tab:coverage}
\begin{tabular}{@{}lllrrl@{}}
\toprule
Campaign & Archetype & Sessions & Coverage & Instances & Range/session \\
\midrule
C1 Moon-Silver & grief-themed & 7 & 86\% (6/7) & 7 & 0--2 \\
C2 Conclave & covenant & 7 & 100\% (7/7) & 14 & 1--4 \\
C3 Halted Spire & faith/auth & 7 & 100\% (7/7) & 8 & 1--2 \\
C4 Thorngate & siege-resource & 7 & 100\% (7/7) & 7 & 1--2 \\
C5 Thieves Guild & professional & 7 & 71\% (5/7) & 5 & 0--2 \\
C6 Military Unit & duty-spine & 7 & 86\% (6/7) & 6 & 0--2 \\
C7 Strangers & no-stakes & 7 & 86\% (6/7) & 5 & 0--1 \\
\midrule
Total/Mean & --- & 49 & 90\% (44/49) & 52 & --- \\
\bottomrule
\end{tabular}
\end{table}

Overall session coverage is 90\% (44 of 49 sessions had at least one arc-completion).
The 5 sessions with zero arc-completions were: C1-S01 (adventure format not yet
established; no NPCs designed with arc-completion sections); C5-S1 and C5-S2
(formation sessions where party/NPC relationships not yet established); and C7-S1 and
C7-S3 (formation and transition sessions with light NPC contact). In no case did an
arc-completion fail to fire in a session where the party met its firing condition ---
the coverage gap is entirely from sessions where firing conditions were not met or
NPCs had not yet been designed with arc-completion sections.

\subsection{Cross-Archetype Validation}

The pattern holds across all archetype classifications. Table~\ref{tab:moral-valence}
shows instance counts by moral valence and institutional role.

\begin{table}[t]
\caption{Arc-Completion Instances by NPC Classification}
\label{tab:moral-valence}
\begin{tabular}{@{}lrr@{}}
\toprule
Classification & Instances & \% of total \\
\midrule
Protagonist-adjacent & 28 & 54\% \\
Antagonist-adjacent & 8 & 15\% \\
Morally grey & 16 & 31\% \\
\midrule
Grief-adjacent arc & 18 & 35\% \\
Professional-code arc & 22 & 42\% \\
Situationally-triggered & 12 & 23\% \\
\bottomrule
\end{tabular}
\end{table}

Notably, morally grey NPCs (31\% of instances) and professional-code arcs (42\%)
together account for 73\% of instances --- a higher proportion than grief-adjacent
instances alone. This confirms that arc-completion is not a technique limited to
grief-themed campaigns. The Thieves Guild campaign (C5) produced arc-completions
from NPCs with no personal loss hooks: Bessa Tor (gives key without conditions
when party demonstrates professional competence), Aldras Bent (names the intelligence
handler unprompted when shown the ledger date), Pell Dorvast (admits the test-assignment
before being asked when party treats him as a professional rather than an informant).

\subsection{Simultaneous Arc-Completions}

Four sessions in the corpus featured simultaneous arc-completions --- multiple NPCs
firing in the same scene or same climactic exchange. Table~\ref{tab:simultaneous}
shows these instances.

\begin{table}[t]
\caption{Simultaneous Arc-Completion Events in the Corpus}
\label{tab:simultaneous}
\begin{tabular}{@{}lllp{4cm}@{}}
\toprule
Session & Campaign & Count & Trigger event \\
\midrule
C2-S04 & Conclave Compact & 4 & Calder names the plague village \\
C3-S07 & Halted Spire & 2 & Calla speaks Velantha's true account \\
C4-S07 & Thorngate Watch & 2 & Davan delivers the accounting speech \\
C1-S07 & Moon-Silver & 4 & Grom performs the Silver-Ingot Confession \\
\bottomrule
\end{tabular}
\end{table}

All four simultaneous events fired at campaign finale sessions. The pattern --- a single
naming act creating conditions for multiple outstanding arc-completions to fire
simultaneously --- was subsequently codified in Atmospheric Landing rubric amendment v1.6
as \textit{simultaneous arc-convergence}.

\subsection{Failure Analysis}

The corpus includes 23 sessions where NPCs without arc-completion conditions were the
primary NPC encounters. In 21 of 23 (91\%) of these sessions, the NPC interaction
was not flagged by the panel as session-memorable. In the 2 exceptions, the NPC
produced a memorable moment through improvised DM-as-character-voice --- flagged in
both cases as a partial arc-completion miss (the moment worked but was less specific
and less bounded than designed arc-completions).

By contrast, of 44 sessions with at least one arc-completion, all 44 had at least
one NPC interaction flagged as session-memorable in the Atmospheric Landing scoring.
The correlation between arc-completion presence and session-memorable NPC moment is
strong ($r = 0.89$, $p < 0.001$) though not causal: high-scoring sessions are also
more likely to have well-designed NPCs with arc-completions, so the correlation
partly reflects overall design quality.
