%% glyph-campaign-spine/main.tex
%% The 7-Session Campaign Spine: Evidence-Based Architecture for
%% Emotionally Complete Narrative Arcs
%%
%% Target venue: ACM CHI 2027
%% Evidence base: Marathon D&D Workshop — 7 complete campaigns, 49 sessions

\documentclass[sigconf]{acmart}

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{microtype}
\usepackage{array}

\acmConference[CHI '27]{CHI Conference on Human Factors in Computing Systems}{May 2027}{Yokohama, Japan}
\acmDOI{10.1145/XXXXXXX.XXXXXXX}
\acmISBN{978-x-xxxx-xxxx-x/27/05}
\copyrightyear{2027}
\acmYear{2027}
\setcopyright{acmlicensed}

\begin{document}

\title{The 7-Session Campaign Spine:\\
       Evidence-Based Architecture for Emotionally Complete Narrative Arcs}

\author{Gio Della-Libera}
\affiliation{%
  \institution{Independent Research}
  \city{Seattle}
  \state{WA}
  \country{USA}
}
\email{giodl@microsoft.com}

\begin{abstract}
Emotional payoffs in tabletop role-playing game (TRPG) campaigns are structurally
unreliable: most long-form campaigns fail to reach a designed conclusion, and those that
do complete often do so through DM improvisation rather than designed architecture.
We present the \textit{campaign spine} --- a five-component operational document pairing
a central question, a 4-hint delivery schedule, branching end-conditions, a
PC-spotlight schedule, and a PC-authored finale deliverable specification --- and report
its validation across 7 complete Dungeons \& Dragons 5e campaigns (49 sessions total)
in the Marathon AI-simulated TRPG workshop. All 7 campaigns completed within 7 sessions
(100\% completion rate). Five of 7 achieved Route D or E, the end-conditions requiring
the highest degree of sustained player engagement with the spine's design. All 28 hints
across all 7 campaigns were delivered (100\%). All 7 PC-authored deliverables were
produced and logged verbatim. Mean sessions-to-first-binding-PASS is 4.5 across all
campaigns. The spine is archetype-independent: it produced equivalent completion and
quality outcomes for grief-themed parties (Campaigns 1--3), siege-resource parties
(Campaign 4), professional-code parties (Campaign 5), duty-spine parties (Campaign 6),
and parties with no designed personal stakes (Campaign 7). We describe each component,
report the cross-campaign evidence, and analyze which structural features most
contribute to reliable finale completion.
\end{abstract}

\keywords{campaign design, narrative arc, hint schedule, route conditions, TRPG,
          emotionally complete narrative, AI simulation, game design methodology,
          interactive storytelling}

\maketitle

\input{sections/01-introduction}
\input{sections/02-related-work}
\input{sections/03-components}
\input{sections/04-evaluation}
\input{sections/05-discussion}
\input{sections/06-conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
