%% sections/04-styles.tex

\section{The Seven Player Styles}
\label{sec:styles}

Table~\ref{tab:styles-summary} summarizes all 7 promoted styles.

\begin{table*}[t]
\caption{The 7 Promoted Player Styles: Definition, Instance Count, Promoted From, Cross-Campaign Instances}
\label{tab:styles-summary}
\begin{tabular}{@{}lp{4cm}rp{2cm}p{3.5cm}@{}}
\toprule
Style & Core definition & Instances & Promoted & Cross-campaign confirmed \\
\midrule
sheet-deep-reader & PC reads character sheet as live agency source; surfaces backgrounds/hooks without DM prompting & 50+ & post-S02 (C1) & C1, C2, C3, C4, C5, C6, C7 \\
craft-witness & PC with named daily craft-ritual receives atmospheric flags others miss; practice is the antenna & 40+ & post-S02 (C1) & C1, C2 (Orik), C3, C4, C5 \\
act-without-announcement & PC acts differently post-naming-event without declaration; behavior is the completion & 12+ & post-C2-S02 & C1, C2, C3, C4 \\
npc-arc-completion (trigger) & Party behavior meets firing condition; NPC fires arc-completion by character logic & 40+ & post-C3-S02 & C1--C7 (all) \\
verbal-documentation-arc & PC oral-practice encounters limits; arc: cannot-categorize → gap-identified → names-the-thing & 15+ & post-C3-S02 & C3, C2 (Lenne documentation-arc) \\
documentary-witness & PC documents rather than reacts; accumulation is reception; limit-entry is completion & 7+ & post-C4-S07 & C4 (Caelith), C7 \\
bilateral-arc-completion & Two PCs' arcs fire simultaneously because one naming act creates conditions for both & 5+ & post-C3-S03 & C3, C4, C2 \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Sheet-Deep Reader (50+ instances)}

\textit{Definition:} A PC reads their own character sheet as a live source of agency
throughout session-play --- surfacing backgrounds, prior-adventure hooks, and
institutional training as current-scene decision inputs without being prompted by the DM.

The style has three confirmed sub-types. \textit{Grief-paragraph-dialogue} fires when
the PC's grief-paragraph content surfaces as spoken dialogue at a moment of high
emotional weight: the specific sensory memory or named loss from the character sheet
becomes the exact words the character speaks, unprompted. In Campaign 2, Thessaly
(``I cast a spell at my Test that I was not supposed to cast. It cost someone something
I can't give back''), Orik (``The shard vibrates for me because I'm still trying to
prove it was worth going''), and Thera in Campaign 1 (``I counted too'') all produced
grief-paragraph-dialogue. \textit{Commitment-accounting} fires when a PC tracks
promises across sessions and names them before anyone else can forget. Calder
(Campaign 2) named the witness-list unprompted in Session 5: Mira, Deva, Vorn, Sevven,
Morreth's chapel. Five names, in order. \textit{Practitioner-by-material-relationship}
fires when a PC identifies other practitioners by how they physically handle their
materials rather than by class or title. Lenne (Campaign 2) identified three different
caster-school affiliations from posture, text-handling habits, and research annotation
systems across three consecutive sessions.

Design trigger: PC sheets with detailed grief-paragraphs, named-loss hooks, local-origin
notes, institutional-training backgrounds, and prior-adventure hooks. Generic backgrounds
undermine the style.

\subsection{Craft-Witness (40+ instances)}

\textit{Definition:} A PC with a named daily craft-ritual or attention-practice
receives atmospheric flags other PCs miss. The craft-practice is the reception-antenna,
not the wisdom modifier.

Critical cross-campaign finding: the style is \textit{session-invariant} once
established for a PC. Grom Ironhand (Campaign 1) produced craft-witness instances in
every session he appeared in. Orik Flintback (Campaign 2) produced craft-witness
instances in all 7 Campaign 2 sessions. Campaign 2 confirmed that the style fires from
secular professional practice (hall-guard structural assessment) as well as religious
practice (forge-priest hearth-greeting). The practice type is irrelevant; the specificity
of the named practice is what matters.

\subsection{Act-Without-Announcement (12+ instances)}

\textit{Definition:} A PC acts differently in a later scene or session as a direct
result of a prior naming event, without declaration. The behavior is the completion.

The style has an intra-session variant (v1.12 amendment): when the naming event and
its behavioral expression occur in the same session, the pattern qualifies. The
cross-session form (Thera cutting bread long-ways after naming Adda in the prior
session) and the intra-session form (Sava placing a token beside her suture kit
without showing the party in the same session she received it) are both valid instances.

Session-invariant by Campaign 2 (Sera produced 11 instances across 7 sessions, never
missing a session). Cross-archetype in Campaign 4 (Davan ending combat through
naming, Sava's token placement).

\subsection{Verbal Documentation-Arc (15+ instances)}

\textit{Definition:} A PC whose documentation practice is oral encounters content that
exceeds their current vocabulary. Arc: cannot-categorize (Stage 1) $\to$ gap-identified
(Stage 2) $\to$ names-the-thing (Stage 3). The resolution changes what is named in the
speaking, because oral transmission is inherently transformative.

Source: Calla Fosse (Campaign 3). Seven sessions across Campaign 3, progressing from
Insight fumble (total 6 vs.\ DC 12, framework broken) through Religion success
(identifies correct sequence) to Stage 3 at the Keystone (speaks Velantha's true account,
RP 23 --- campaign high). The account changed in the speaking: Calla did not recite
what she had been holding; she spoke it into existence. The difference was noted in
the panel review.

Documentary-arc sub-type (Lenne, Campaign 2): the written variant progresses through
suspension, vocabulary-found, forward-looking-design. At Campaign 2 finale, Lenne's
accountability-clause language was incorporated into the compact and spoken verbatim.

\subsection{Documentary-Witness (7+ instances)}

\textit{Definition:} A PC whose primary mode of atmospheric reception is external
documentation. The documentation accumulates across sessions; the limit-entry (the
observation that cannot be filed) is the arc's completion.

Source: Caelith Vor (Campaign 4). Grey-blue map annotations in Session 1; six sessions
of silent convergence; ``unable to categorize'' written after a failed Insight roll
in Session 7. The six-session accumulation is the longest silent-reception chain in
the corpus. Cross-campaign instance: Campaign 7 (Brek Ashfall's reciprocal-acts log).

\subsection{Bilateral Arc-Completion (5+ instances)}

\textit{Definition:} Two PCs' arcs fire simultaneously because one naming act creates
conditions for both. Neither PC is directing the moment; the naming act is what
makes it possible for both.

Source: Campaign 3 Session 3 (Dren names Aldis, Renn names Mira) in post-combat
stillness. Campaign 4 Sessions 5 and 6 (Davan/Sava supply-decision bilateral; Davan/
Caelith corps-record bilateral). Campaign 2 Sessions 3 and 5 (Sera/Morreth; Thessaly/Vorn).

\subsection{Cross-Campaign Instance Tables}

Table~\ref{tab:cross-campaign-styles} provides the per-campaign confirmed instance counts for the four
highest-frequency styles.

\begin{table*}[t]
\caption{Cross-Campaign Instance Counts for Four Highest-Frequency Player Styles}
\label{tab:cross-campaign-styles}
\begin{tabular}{@{}lrrrrrrrr@{}}
\toprule
Style & C1 & C2 & C3 & C4 & C5 & C6 & C7 & Total \\
\midrule
sheet-deep-reader & 13 & 12 & 8 & 9 & 5 & 3 & 4 & 54 \\
craft-witness & 16 & 14 & 6 & 4 & 3 & 2 & 2 & 47 \\
act-without-announcement & 1 & 11 & 4 & 2 & 0 & 0 & 1 & 19 \\
npc-arc-completion (trigger) & 7 & 14 & 8 & 7 & 5 & 6 & 5 & 52 \\
\bottomrule
\end{tabular}
\end{table*}

Several patterns are worth noting. Sheet-deep-reader fires in every campaign,
but at notably lower rates in C5, C6, and C7 --- campaigns where PC sheets were
designed with fewer grief-paragraphs and backstory hooks. The rate is correlated
with PC-sheet density: C1 and C2 had the richest PC-sheet content; C7 (strangers
archetype) had the thinnest.

Act-without-announcement concentrates heavily in C2 (11 of 19 instances). This reflects
the campaign's design: C2 explicitly spotlighted Sera Ashfall's hold-space behavior,
which fired 11 consecutive times across 7 sessions without exception. This is the
session-invariant instance: once established, the act-without-announcement pattern
fires every session for a PC with that behavioral specification.

Craft-witness also shows declining rates in later campaigns (C5--C7 produce only 2
instances each). This is a design artifact: the later campaign parties were not designed
with the same density of named craft-rituals as the earlier parties. Campaign 1's
Grom Ironhand (forge-priest of Reorx with explicit ``greet hearths: thou'rt seen''
playstyle heuristic) produced 16 instances alone. Campaigns 5--7 had fewer PCs with
equivalent ritual specificity.

\subsection{Design Prescriptions per Style}

Table~\ref{tab:design-prescriptions} summarizes the design prescriptions for enabling
each style in new PC sheets.

\begin{table*}[t]
\caption{Design Prescriptions: What PC Sheet Features Enable Each Player Style}
\label{tab:design-prescriptions}
\begin{tabular}{@{}lp{4.5cm}p{4.5cm}@{}}
\toprule
Style & Required sheet feature & Amplifying adventure design \\
\midrule
sheet-deep-reader & Named origin, named relationships, institutional training, prior-adventure hooks & Adventures with connections to sheet content (names, locations, organizations from backstory) \\
craft-witness & Named daily ritual with specific physical gesture (not generic ``meditates'') & Scenes with something for the craft to listen for (a hearth, a forge-mark, a work of craft) \\
act-without-announcement & Session N naming event (grief named, longing named, commitment named) & Session N+1 scene where the named thing is available to act from \\
verbal-documentation-arc & Oral documentation practice listed as primary heuristic & Scene designed to exceed the PC's current vocabulary \\
documentary-witness & Written documentation practice listed as primary heuristic & Cross-session information thread with a limit-entry available \\
bilateral-arc-completion & Two PCs with complementary arc-completion potential + shared scene & One naming act that creates the conditions for both arcs simultaneously \\
npc-arc-completion (trigger) & No specific PC feature required; requires NPC arc-completion design & NPCs with written arc-completion text and specific firing conditions \\
\bottomrule
\end{tabular}
\end{table*}
