%% sections/05-deliverables.tex

\section{PC-Authored Finale Deliverables}
\label{sec:deliverables}

Across 4 campaigns, a PC-authored deliverable --- a speech, document, forgery, or act ---
was constructed across multiple sessions and deployed verbatim at the campaign finale.
Table~\ref{tab:deliverables} summarizes these deliverables.

\begin{table*}[t]
\caption{PC-Authored Finale Deliverables: Campaign, PC, Type, Build Sessions, Key Design Feature}
\label{tab:deliverables}
\begin{tabular}{@{}llllrp{3.5cm}@{}}
\toprule
Campaign & PC & Deliverable Type & Build sessions & Words (approx.) & Key sheet feature \\
\midrule
C1 Moon-Silver & Grom Ironhand & Spoken confession (Silver-Ingot) & S03-S07 (5 sessions) & 280 & Reorx-craft heuristic + grief paragraph (Duren) \\
C2 Conclave & Thessaly & Compact opening sentence & C2-S01 to C2-S07 (7 sessions) & 45 & Named Test wound; decision-order heuristic \\
C4 Thorngate Watch & Davan Fostermark & Accounting speech & C4-S04 to C4-S07 (4 sessions) & 480 & Honor-accounting heuristic; Orel grief paragraph \\
C5 Thieves Guild & Mira Calloway & False dossier (MARTEN-7) & C5-S01 to C5-S07 (7 sessions) & 120 & Framework-accumulation heuristic; professional code \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Construction Pattern}

All four deliverables show the same construction pattern: the deliverable begins as a
structural invitation in the PC's character sheet (a grief paragraph, a professional
code, a decision-order heuristic) and accumulates substance across sessions as the
campaign's events provide material. The deliverable is not written by the designer;
it is grown by the system playing the PC in response to campaign events.

\paragraph{Grom's Silver-Ingot Confession.}
The confession began with Duren Coalbeard's grief paragraph (PC sheet) and accumulated
across 5 sessions of Reorx-related events: the re-interment of Duren's hammer (S04),
the three-vein synthesis (S06), and the workshop approach (S07). At the forge, Grom
named two shames: Aelwen's (things made to hold grief; the withheld tool) and his own
(``I was slow and I was proud and I called it patience''). Neither was scripted by the
module; both were shaped by session events acting on the PC sheet's Reorx-forge-craft
identity.

\paragraph{Davan's Accounting Speech.}
The speech accumulated across four sessions of Oath-collision events. Session 4 (Oath
cannot tell me this); Session 5 (institutional question as architecture); Session 6
(cost of correctness); Session 7 synthesis (Orel named). The sheet's honor-accounting
heuristic (``name what was spent, by whom, without advocacy'') was the structural
constraint that gave the speech its form. The content --- including the naming of
Orel --- emerged from session events. The speech is logged verbatim in the session
handoff at 480 words.

\paragraph{Mira's False Dossier.}
The dossier (MARTEN-7) accumulated across 7 sessions as Mira's framework-accumulation
heuristic (building evidence without surfacing it) gathered 6 components. At the finale,
she deployed the completed dossier: ``Pale described party as low-yield fencing network,
no intelligence activity, RI value: none, recommend inactive.'' The dossier neutralized
the campaign's antagonist through professional action rather than confrontation --- a
Route B outcome consistent with the sheet's professional-code identity.

\subsection{Design Implication}

PC-authored deliverables require two design features: (a) a sheet feature that provides
the structural constraint (what kind of deliverable the PC is capable of producing and
why) and (b) a campaign spine that provides enough material-events to load the
deliverable across sessions. Without both, the deliverable either does not emerge or
emerges as a thin, underloaded product.

The marathon campaign spine explicitly designates a PC-authored deliverable for each
campaign and builds material-events across sessions specifically to load it. The result
is a deliverable that feels like an inevitable conclusion rather than a scripted speech ---
because the PC's sheet and the campaign's events accumulated toward it together, not
because the designer wrote it in advance.
